k-plus Anticlustering: An Improved k-means Criterion for Maximizing Between-Group Similarity

### 2023-05-07 paper.Rmd
typos

[^dissimilarityfromfeatures]: It is still possible to convert the attribute data into a dissimilarity matrix, e.g. by computing all pairwise Euclidean distances, and then optimize a distance-based anticlustering criterion to obtain a partitioning. For likert scale data or binary attributes, the Manhattan distance may be preferred over the Euclidean distance; for mixed-type attributes, it is even possible to employ a combined distance metric [@rokach2005].

@@

The weights $w_1$ and $w_2$ can be chosen to adjust the relative importance of equalizing means and variances, respectively. By maximizing $\mathit{SSE_{kplus}}$ we seek to obtain a partitioning where means as well as variances are similar between groups. However, research on multiobjective optimization indicates that solutions oftentimes fail to satisfactorily address several criteria if the underlying objectives are in conflict rather than in alignment (Ferligoj & Batagelj, 1992). As $\mathit{SSE_{kplus}}$ is a novel criterion, it is so far unknown if the two constituting criteria are aligned. That is, it has yet to be demonstrated that maximizing $\mathit{SSE_{kplus}}$ can yield a partition where means as well as variances are similar between groups. A motivating example illustrates that aiming to simultaneously optimize similarity of means and variances may be feasible. I created 14 data points following a univariate normal distribution ($M = 15$, $SD = 3$) and then generated all `r n_partitions(14, 2)` possible ways to partition the 14 data points into $K = 2$ equal-sized groups. Figure 1 shows the scatterplot of the criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ across partitions. First, it is shown that the criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ seem to be rather well aligned: There is a high concentration of partitions where both criteria have high values. At the same time, there are many partitions that fulfil only one of the criteria very well, but fail to also fulfil the other criterion. The partition that is represented by the squared symbol has the overall best highest $\mathit{SSE}$, but is outperformed by many other partitions with regard to $\mathit{SSE_{Var}}$. Similarly, the partition that maximizes $\mathit{SSE_{Var}}$ (represented by the triangular symbol) is not well-suited to optimize $\mathit{SSE}$ at the same time. Thus, when only considering one of the objectives, the mathematically optimal partitioning fails to establish groups that are similar both with regard to their means and variances. The diamond symbol represents the partition that maximizes the unweighted sum of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$, which yields `r df$kmeans[which.max(df$kplus)] / max(df$kmeans) * 100`% of the global maximum $\mathit{SSE}$, and `r df$kvar[which.max(df$kplus)] / max(df$kvar) * 100`% of the global maximum value of $\mathit{SSE_{Var}}$. Thus, even using a very simple unweighted sum approach, the two constituting criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ can be fulfilled to an astonishing degree.

@@

After the four anticlustering methods and the random allocation procedure had been applied to the 10,000 data sets for $K = 2$, $K = 3$, $K = 4$, I investigated how well the competing methods were able to create similar groups. To this end, I analyzed the discrepancy in means, standard deviations, skewness and kurtosis between groups---for each data set for each $K$. To quantify the discrepancy in group means, the following computation was used: for each of the (2, 3, 4 or 5) features, all (2, 3 or 4) group means were computed. For each feature, the difference between the minimum and maximum group mean was used as a measure of discrepancy; the global discrepancy in means ($\Delta M$) was then determined as the average discrepancy across features. The same procedure was applied to compute the discrepancy in standard deviations ($\Delta \mathit{SD}$), skewness ($\Delta \mathit{Skew}$) and kurtosis ($\Delta \mathit{Kurtosis}$).

@@

Figure 2 shows that with increasing $N$, k-means and the k-plus methods converged on the same level with regard to minimizing $\Delta \mathit{M}$. This is an important observation because it shows that k-plus anticlustering is capable of addressing multiple objectives at the same time---especially when $N$ increases. However, while k-plus anticlustering was capable of addressing multiple objectives simultaneously, increasing the number of optimization criteria made it more difficult to fulfill each single criterion: generalized k-plus performed worse at minimizing $\Delta \mathit{M}$ and $\Delta \mathit{SD}$ than standard k-plus that only considered the mean and variances. However, as intended, generalized k-plus was best at minimizing $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$ and at the same time, this k-plus objective maintained a comparably good level of addressing $\Delta \mathit{M}$ and $\Delta \mathit{SD}$, outperforming diversity anticlustering.

@@

In the first example, I used the OASIS norming data that is freely available online [@kurdi2017introducing]. Kurdi et al. assembled 900 open-access color images and collected ratings on two affective dimensions: arousal and valence. @brielmann2019intense measured how the same 900 images were rated with regard to beauty. I used the function `kplus_anticlustering()` to divide all images into 9 groups of 100 images each. The first argument of the function call is a variable named `features`, which contains the image data as a table: 900 rows representing images and 3 columns representing the features beauty, arousal, and valence. An OSF repository accompanying this manuscript contains the full code to reproduce the example [@osf2023].

@@

Two simulation studies and examples on real norming data showed that the k-plus criterion is well-suited to maximize between-group similarity with regard to multiple criteria. Given that similarity with regard to variances was achieved with almost no sacrifice to similarity in means, k-plus anticlustering should be considered as the default replacement of classical k-means when striving for between-group similarity.[^binarydata] The results did however show that fulfilling multiple criteria at the same time can be challenging in small data sets. In practice, users can try out different objectives and see which specification fulfills their needs. A reasonable approach is to start "greedy", e.g. by specifying that multiple distribution moments should equalized. If the results of such a greedy setup are not satisfying, users can reduce their requirements by opting for fewer criteria (e.g., only the mean and variance). Using the open source R package `anticlust`, the (repeated) application of k-plus anticlustering is easily accessible and free.

@@

The present paper focused on the new k-plus criterion for anticlustering, but only briefly considered the multicriterion nature of this criterion. A simple unweighted sum approach was used to optimize the k-plus objective, while more sophisticated algorithmic procedures exist to tackle multicriterion optimization problems [e.g., @brusco2009cross]. The sole focus on the objective function followed two reasonings. First, the applications and the simulations showed that the unweighted standardization approach yielded satisfying results that probably suffice in most applications. Moreover, it is possible that k-plus anticlustering has a fortunate solution space where many suitable partitions exist that fulfill multiple criteria to a satisfactory degree (as suggested by Figure 1). Still, more sophisticated approaches may yield theoretically improved results, as well as the opportunity to deliberately select among partitionings that may be preferred according to different k-plus objectives. @brusco2019 gave an example of how direct algorithms can be employed to optimize multiple criteria in anticlustering applications aiming for high within-group heterogeneity. Future research should investigate if k-plus anticlustering may profit from similar algorithmic treatment. Specifically, it is of high interest to investigate if the bicriterion approach by Brusco et al. can be transformed into a multicriterion algorithm for the k-plus objective. A second reason for focusing on the k-plus objective per se---independently from the algorithm that is used to optimize it---was to obtain a fair comparison to existing anticlustering objectives. In the simulation study, differences in performance could not have been interpreted if the objective function as well as the algorithm used to optimize it had differed between anticlustering methods. It was of particular interest to compare the k-plus objective with the popular diversity objective. @papenberg2020 compared k-means anticlustering and diversity anticlustering[^clusterediting] in a simulation study and concluded that the diversity criterion should be preferred because k-means does not equalize the spread of the distribution between groups. In contrast, maximizing the diversity leads to an appropriate balance between location and spread. Using the k-plus extension, however, the k-means criterion has regained attractivity: When opting for similarity in means and variances---and for increasing $N$, skewness and kurtosis as well---, k-plus anticlustering clearly outperformed diversity anticlustering. Future research should investigate whether this advantage also holds for other data distributions than normal. For example, it may be hypothesized that the generalized k-plus criterion is even more useful when applied to data that is not symmetric or contains outliers, because it is able to explicitly incorporate the skewness and kurtosis as criteria in the anticlustering process.

>>>>

[^dissimilarityfromfeatures]: It is still possible to convert the attribute data into a dissimilarity matrix, e.g. by computing all pairwise Euclidean distances, and then optimize a distance-based anticlustering criterion to obtain a partitioning. For Likert scale data or binary attributes, the Manhattan distance may be preferred over the Euclidean distance; for mixed-type attributes, it is even possible to employ a combined distance metric [@rokach2005].

@@

The weights $w_1$ and $w_2$ can be chosen to adjust the relative importance of means and variances in the anticlustering process. By maximizing $\mathit{SSE_{kplus}}$ we seek to obtain a partitioning where means as well as variances are similar between groups. However, research on multiobjective optimization indicates that solutions oftentimes fail to satisfactorily address several criteria if the underlying objectives are in conflict rather than in alignment (@ferligoj1992direct). As $\mathit{SSE_{kplus}}$ is a novel criterion, it is so far unknown if the two constituting criteria are aligned. That is, it has yet to be demonstrated that maximizing $\mathit{SSE_{kplus}}$ can yield a partition where means as well as variances are similar between groups. A motivating example illustrates that aiming to simultaneously optimize similarity of means and variances may be feasible. I created 14 data points following a univariate normal distribution ($M = 15$, $SD = 3$) and then generated all `r n_partitions(14, 2)` possible ways to partition the 14 data points into $K = 2$ equal-sized groups. Figure 1 shows the scatter plot of the criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ across partitions. First, it is shown that the criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ seem to be rather well aligned: There is a high concentration of partitions where both criteria have high values. At the same time, there are many partitions that fulfil only one of the criteria very well, but fail to also fulfil the other criterion. The partition that is represented by the squared symbol has the overall best highest $\mathit{SSE}$, but is outperformed by many other partitions with regard to $\mathit{SSE_{Var}}$. Similarly, the partition that maximizes $\mathit{SSE_{Var}}$ (represented by the triangular symbol) is not well-suited to optimize $\mathit{SSE}$ at the same time. Thus, when only considering one of the objectives, the mathematically optimal partitioning fails to establish groups that are similar both with regard to their means and variances. The diamond symbol represents the partition that maximizes the unweighted sum of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$, which yields `r df$kmeans[which.max(df$kplus)] / max(df$kmeans) * 100`% of the global maximum $\mathit{SSE}$, and `r df$kvar[which.max(df$kplus)] / max(df$kvar) * 100`% of the global maximum value of $\mathit{SSE_{Var}}$. Thus, even using a very simple unweighted sum approach, the two constituting criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ can be fulfilled to an astonishing degree.

@@

After the four anticlustering methods and the random allocation procedure had been applied to the 10,000 data sets for $K = 2$, $K = 3$, $K = 4$, I investigated how well the competing methods were able to create similar groups. To this end, I analysed the discrepancy in means, standard deviations, skewness and kurtosis between groups---for each data set for each $K$. To quantify the discrepancy in group means, the following computation was used: for each of the (2, 3, 4 or 5) features, all (2, 3 or 4) group means were computed. For each feature, the difference between the minimum and maximum group mean was used as a measure of discrepancy; the global discrepancy in means ($\Delta M$) was then determined as the average discrepancy across features. The same procedure was applied to compute the discrepancy in standard deviations ($\Delta \mathit{SD}$), skewness ($\Delta \mathit{Skew}$) and kurtosis ($\Delta \mathit{Kurtosis}$).

@@

Figure 2 shows that with increasing $N$, k-means and the k-plus methods converged on the same level with regard to minimizing $\Delta \mathit{M}$. This is an important observation because it shows that k-plus anticlustering is capable of addressing multiple objectives at the same time---especially when $N$ increases. However, while k-plus anticlustering was capable of addressing multiple objectives simultaneously, increasing the number of optimization criteria made it more difficult to fulfil each single criterion: generalized k-plus performed worse at minimizing $\Delta \mathit{M}$ and $\Delta \mathit{SD}$ than standard k-plus that only considered the mean and variances. However, as intended, generalized k-plus was best at minimizing $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$ and at the same time, this k-plus objective maintained a comparably good level of addressing $\Delta \mathit{M}$ and $\Delta \mathit{SD}$, outperforming diversity anticlustering.

@@

In the first example, I used the OASIS norming data that is freely available online [@kurdi2017introducing]. Kurdi et al. assembled 900 open-access colour images and collected ratings on two affective dimensions: arousal and valence. @brielmann2019intense measured how the same 900 images were rated with regard to beauty. I used the function `kplus_anticlustering()` to divide all images into 9 groups of 100 images each. The first argument of the function call is a variable named `features`, which contains the image data as a table: 900 rows representing images and 3 columns representing the features beauty, arousal, and valence. An OSF repository accompanying this manuscript contains the full code to reproduce the example [@osf2023].

@@

Two simulation studies and examples on real norming data showed that the k-plus criterion is well-suited to maximize between-group similarity with regard to multiple criteria. Given that similarity with regard to variances was achieved with almost no sacrifice to similarity in means, k-plus anticlustering should be considered as the default replacement of classical k-means when striving for between-group similarity.[^binarydata] The results did however show that fulfilling multiple criteria at the same time can be challenging in small data sets. In practice, users can try out different objectives and see which specification fulfils their needs. A reasonable approach is to start "greedy", e.g. by specifying that multiple distribution moments should equalized. If the results of such a greedy setup are not satisfying, users can reduce their requirements by opting for fewer criteria (e.g., only the mean and variance). Using the open source R package `anticlust`, the (repeated) application of k-plus anticlustering is easily accessible and free.

@@

The present paper focused on the new k-plus criterion for anticlustering, but only briefly considered the multicriterion nature of this criterion. A simple unweighted sum approach was used to optimize the k-plus objective, while more sophisticated algorithmic procedures exist to tackle multicriterion optimization problems [e.g., @brusco2009cross]. The sole focus on the objective function followed two reasonings. First, the applications and the simulations showed that the unweighted standardization approach yielded satisfying results that probably suffice in most applications. Moreover, it is possible that k-plus anticlustering has a fortunate solution space where many suitable partitions exist that fulfil multiple criteria to a satisfactory degree (as suggested by Figure 1). Still, more sophisticated approaches may yield theoretically improved results, as well as the opportunity to deliberately select among partitions that may be preferred according to different k-plus objectives. @brusco2019 gave an example of how direct algorithms can be employed to optimize multiple criteria in anticlustering applications aiming for high within-group heterogeneity. Future research should investigate if k-plus anticlustering may profit from similar algorithmic treatment. Specifically, it is of high interest to investigate if the bicriterion approach by Brusco et al. can be transformed into a multicriterion algorithm for the k-plus objective. A second reason for focusing on the k-plus objective per se---independently from the algorithm that is used to optimize it---was to obtain a fair comparison to existing anticlustering objectives. In the simulation study, differences in performance could not have been interpreted if the objective function as well as the algorithm used to optimize it had differed between anticlustering methods. It was of particular interest to compare the k-plus objective with the popular diversity objective. @papenberg2020 compared k-means anticlustering and diversity anticlustering[^clusterediting] in a simulation study and concluded that the diversity criterion should be preferred because k-means does not equalize the spread of the distribution between groups. In contrast, maximizing the diversity leads to an appropriate balance between location and spread. Using the k-plus extension, however, the k-means criterion has regained attractivity: When opting for similarity in means and variances---and for increasing $N$, skewness and kurtosis as well---, k-plus anticlustering clearly outperformed diversity anticlustering. Future research should investigate whether this advantage also holds for other data distributions than normal. For example, it may be hypothesized that the generalized k-plus criterion is even more useful when applied to data that is not symmetric or contains outliers, because it is able to explicitly incorporate the skewness and kurtosis as criteria in the anticlustering process.
